\vspace{1.0cm} \Large \bf Integrating influenza antigenic dynamics with molecular evolution

### 2013-07-23 flux.tex
Notation change: drift term from \beta to \mu to agree with standard diffusion notation and also to avoid confusion with incidence regression terms.

% notation
\usepackage{amsmath}
\usepackage{amssymb}
\newcommand{\virus}{\mathbf{x}}						% virus coordinate
\newcommand{\serum}{\mathbf{y}}						% serum coordinate
\newcommand{\viruses}{\mathbf{X}}					% set of virus coordinates
\newcommand{\sera}{\mathbf{Y}}						% set of serum coordinates
\newcommand{\ve}{v}									% virus effect
\newcommand{\se}{s}									% serum effect
\newcommand{\ves}{\mathbf{v}}						% set of virus effects
\newcommand{\ses}{\mathbf{s}}						% set of serum effects
\newcommand{\point}{f_{\scriptscriptstyle \vert}}	% point likelihood
\newcommand{\threshold}{f_{\textstyle \lrcorner}}	% threshold likelihood
\newcommand{\interval}{f_{\sqcup}}					% interval likelihood
\newcommand{\mdssd}{\varphi}						% MDS standard deviation
\newcommand{\virussd}{\sigma_x}						% virus / diffusion standard deviation
\newcommand{\serumsd}{\sigma_y}						% serum standard deviation
\newcommand{\tree}{\tau}							% phylogeny
\newcommand{\vn}{n}									% number of viruses
\newcommand{\sn}{k}									% number of sera
\newcommand{\normal}{\mathcal{N}}					% normal distribution
\newcommand{\bwithin}{\beta_w}		% within clade drift coefficient
\newcommand{\bsister}{\beta_s}		% sister clade drift coefficient
\newcommand{\bother}{\beta_t}			% across clade drift coefficient
\newcommand{\incclade}[1]{y_\mathrm{#1}}
\newcommand{\driftclade}[1]{x_\mathrm{#1}}
\setlength{\arraycolsep}{2pt}
\newcommand{\smalltwomatrix}[2]{\scriptsize \Big( \begin{matrix} #1 \\ #2 \end{matrix} \Big)}				% pretty inline matrix 
\newcommand{\smallfourmatrix}[4]{\scriptsize \Big( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \Big)}	% pretty inline matrix 
\newcommand{\twomatrix}[2]{\left( \begin{matrix} #1 \\ #2 \end{matrix} \right)}								% pretty inline matrix 
\newcommand{\fourmatrix}[4]{\left( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \right)}					% pretty inline matrix

@@

As presented, multiple configurations of virus and serum locations $\viruses$ and $\sera$ will give the same likelihood of an observed data matrix $\mathbf{H}$.
An example of this phenomenon is shown in Figure~\ref{schematic_map}.
In this case, it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar (Figure~\ref{schematic_map}A) or antigenically divergent (Figure~\ref{schematic_map}B). 
This presents an issue of model identifiability, where absolute, as opposed to relative, antigenic locations cannot be determined from observing the serological data alone.
Thus, in order to achieve more a interpretable model we impose a weak prior on global locations.
In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
To capture this, we replace our previous diffuse prior with an informed prior in which the expected location of viruses and sera increases with date of sampling along dimension one, and each virus and serum location follows an independent normal distribution centered around this temporal expectation, so that
\begin{eqnarray}
	x_{i1} &\sim& \beta \, t_i + \normal(0, \virussd^2) \nonumber \\
	y_{j1} &\sim& \beta \, t_j + \normal(0, \serumsd^2),
\end{eqnarray}
where $t$ is the difference between the date of the indexed virus or serum and the date of the earliest sampled virus or serum, and other dimensions follow $x_{im} \sim \normal(0, \virussd^2)$ and $y_{jm} \sim \normal(0, \serumsd^2)$ for $m\ge2$.
Thus, this model assumes that virus and serum locations drift in a line across the antigenic map at rate $\beta$.
The parameter $\virussd$ determines the breadth of the cloud of virus locations at each point in time, while $\serumsd$ determines the breadth of the cloud of serum locations.

@@

We simultaneously model antigenic locations and genetic relatedness by assuming that virus locations are influenced by evolution following a Brownian motion process \cite{Lemey10}.
To do this, we replace the previous prior specifying independent virus locations with a prior that incorporates covariance based on shared evolutionary history
\begin{equation} \label{ebpprior}
	\viruses \sim \left( \begin{matrix} \beta \, t_1 & 0 \\ \vdots & \vdots \\ \beta \, t_n & 0 \end{matrix} \right) + \, \mbox{Evolutionary Brownian Process}(\virussd, \tree)
\end{equation}
for $P=2$, where $\virussd$ is the volatility parameter of the Brownian motion over virus locations and $\tree$ is a phylogeny specifying tree topology and branch lengths.
Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
In the evolutionary Brownian process, the tips of the phylogeny $\tree$ correspond to the set of virus locations $(\virus_1, \ldots, \virus_\vn)$, and the probability of observing tip locations depends on the locations of internal nodes $(\virus_{n+1}, \ldots, \virus_{2\vn-2})$ and on the location of the root node $\virus_{2\vn-1}$.
This process assumes that a virus location $\virus_i$ follows from the location of its parent virus $\virus_{f(i)}$, and with the addition of drift along dimension 1, is distributed as
\begin{equation} 
	\virus_i \sim (\beta \, d_i, \, 0)^{\prime} + \normal(\virus_{f(i)}, d_i \, \boldsymbol{\Sigma})
\end{equation}
for $P=2$, where $f(i)$ is a function that maps nodes to parental nodes, $d_i$ is the length of the branch connecting virus $i$ to parent virus $f(i)$, and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $\virussd^2$.
The root virus location $\virus_{2\vn-1}$ is assumed to follow a normal distribution with expectation $(\beta \, t_{2\vn-1}, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
The probability of virus locations $p(\viruses | \beta, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
This formulation corresponds to a Wiener process with drift, in which the drift term $\beta$ only influences the expected states of nodes along the phylogeny, but does not influence the covariance structure among these nodes, which remains the same as it does in a standard Wiener process \cite{BrownianMotionHandbook}.
This allows the separation in equation \ref{ebpprior} between drift terms affecting only expectations and the evolutionary Brownian process that includes covariance among virus locations $\virus_1,\ldots,\virus_n$.

@@

Top-level priors for $1/\mdssd^2$, $\beta$, $1/\virussd^2$, and $1/\serumsd^2$ are assumed to follow diffuse $\mbox{Gamma}(a, b)$ distributions  with $a=0.001$ and $b=0.001$.
Under the full model, the posterior probability of observing virus and serum locations given immunological data is factored
\begin{equation}
	p(\viruses,\sera | \mathbf{H}) \propto p(\mathbf{H} | \viruses, \sera, \ses, \ves, \mdssd) \; 
	p(\viruses | \beta, \virussd, \tree) \;
	p(\sera | \beta, \serumsd) \; 
	p(\ses, \ves, \mdssd, \beta, \virussd, \serumsd, \tree).
\end{equation}
We sample from this posterior distribution using the MCMC procedures implemented in the software package BEAST \cite{BEAST17}.
Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus effects $\ve_i$ and serum effects $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\beta$, $\virussd$ and $\serumsd$.
For the present analysis, a two-step approach was taken to sample phylogenies, where a posterior sample of phylogenies was gathered using sequence data and then, in the cartographic analysis, trees from this set were randomly proposed and accepted following the Metropolis-Hastings algorithm \cite{Pagel04}.

@@

MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\beta$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd^2$ and phylogenetic tree $\tree$.
MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.
In all cases when drift parameter $\beta$ was included the MCMC chain mixed well and arrived at the same estimated posterior distribution from different starting points.
However, without drift parameter $\beta$, maps for A/H3N2 showed some degree of metastability, where some chains would converge on one solution and other chains would converge on a different solution.
We favor models that include $\beta$, because its inclusion, in addition to correcting most identifiability issues, yields much improved mixing of antigenic locations.

>>>>

% notation
\usepackage{amsmath}
\usepackage{amssymb}
\newcommand{\virus}{\mathbf{x}}						% virus coordinate
\newcommand{\serum}{\mathbf{y}}						% serum coordinate
\newcommand{\viruses}{\mathbf{X}}					% set of virus coordinates
\newcommand{\sera}{\mathbf{Y}}						% set of serum coordinates
\newcommand{\ve}{v}									% virus effect
\newcommand{\se}{s}									% serum effect
\newcommand{\ves}{\mathbf{v}}						% set of virus effects
\newcommand{\ses}{\mathbf{s}}						% set of serum effects
\newcommand{\point}{f_{\scriptscriptstyle \vert}}	% point likelihood
\newcommand{\threshold}{f_{\textstyle \lrcorner}}	% threshold likelihood
\newcommand{\interval}{f_{\sqcup}}					% interval likelihood
\newcommand{\mdssd}{\varphi}						% MDS standard deviation
\newcommand{\virussd}{\sigma_x}						% virus / diffusion standard deviation
\newcommand{\serumsd}{\sigma_y}						% serum standard deviation
\newcommand{\drift}{\mu}							% drift / advection
\newcommand{\tree}{\tau}							% phylogeny
\newcommand{\vn}{n}									% number of viruses
\newcommand{\sn}{k}									% number of sera
\newcommand{\normal}{\mathcal{N}}					% normal distribution
\newcommand{\bwithin}{\beta_w}		% within clade drift coefficient
\newcommand{\bsister}{\beta_s}		% sister clade drift coefficient
\newcommand{\bother}{\beta_t}			% across clade drift coefficient
\newcommand{\incclade}[1]{y_\mathrm{#1}}
\newcommand{\driftclade}[1]{x_\mathrm{#1}}
\setlength{\arraycolsep}{2pt}
\newcommand{\smalltwomatrix}[2]{\scriptsize \Big( \begin{matrix} #1 \\ #2 \end{matrix} \Big)}				% pretty inline matrix 
\newcommand{\smallfourmatrix}[4]{\scriptsize \Big( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \Big)}	% pretty inline matrix 
\newcommand{\twomatrix}[2]{\left( \begin{matrix} #1 \\ #2 \end{matrix} \right)}								% pretty inline matrix 
\newcommand{\fourmatrix}[4]{\left( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \right)}					% pretty inline matrix

@@

As presented, multiple configurations of virus and serum locations $\viruses$ and $\sera$ will give the same likelihood of an observed data matrix $\mathbf{H}$.
An example of this phenomenon is shown in Figure~\ref{schematic_map}.
In this case, it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar (Figure~\ref{schematic_map}A) or antigenically divergent (Figure~\ref{schematic_map}B). 
This presents an issue of model identifiability, where absolute, as opposed to relative, antigenic locations cannot be determined from observing the serological data alone.
Thus, in order to achieve more a interpretable model we impose a weak prior on global locations.
In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
To capture this, we replace our previous diffuse prior with an informed prior in which the expected location of viruses and sera increases with date of sampling along dimension one, and each virus and serum location follows an independent normal distribution centered around this temporal expectation, so that
\begin{eqnarray}
	x_{i1} &\sim& \drift \, t_i + \normal(0, \virussd^2) \nonumber \\
	y_{j1} &\sim& \drift \, t_j + \normal(0, \serumsd^2),
\end{eqnarray}
where $t$ is the difference between the date of the indexed virus or serum and the date of the earliest sampled virus or serum, and other dimensions follow $x_{im} \sim \normal(0, \virussd^2)$ and $y_{jm} \sim \normal(0, \serumsd^2)$ for $m\ge2$.
Thus, this model assumes that virus and serum locations drift in a line across the antigenic map at rate $\drift$.
The parameter $\virussd$ determines the breadth of the cloud of virus locations at each point in time, while $\serumsd$ determines the breadth of the cloud of serum locations.

@@

We simultaneously model antigenic locations and genetic relatedness by assuming that virus locations are influenced by evolution following a Brownian motion process \cite{Lemey10}.
To do this, we replace the previous prior specifying independent virus locations with a prior that incorporates covariance based on shared evolutionary history
\begin{equation} \label{ebpprior}
	\viruses \sim \left( \begin{matrix} \drift \, t_1 & 0 \\ \vdots & \vdots \\ \drift \, t_n & 0 \end{matrix} \right) + \, \mbox{Evolutionary Brownian Process}(\virussd, \tree)
\end{equation}
for $P=2$, where $\virussd$ is the volatility parameter of the Brownian motion over virus locations and $\tree$ is a phylogeny specifying tree topology and branch lengths.
Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
In the evolutionary Brownian process, the tips of the phylogeny $\tree$ correspond to the set of virus locations $(\virus_1, \ldots, \virus_\vn)$, and the probability of observing tip locations depends on the locations of internal nodes $(\virus_{n+1}, \ldots, \virus_{2\vn-2})$ and on the location of the root node $\virus_{2\vn-1}$.
This process assumes that a virus location $\virus_i$ follows from the location of its parent virus $\virus_{f(i)}$, and with the addition of drift along dimension 1, is distributed as
\begin{equation} 
	\virus_i \sim (\drift \, d_i, \, 0)^{\prime} + \normal(\virus_{f(i)}, d_i \, \boldsymbol{\Sigma})
\end{equation}
for $P=2$, where $f(i)$ is a function that maps nodes to parental nodes, $d_i$ is the length of the branch connecting virus $i$ to parent virus $f(i)$, and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $\virussd^2$.
The root virus location $\virus_{2\vn-1}$ is assumed to follow a normal distribution with expectation $(\drift \, t_{2\vn-1}, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
The probability of virus locations $p(\viruses | \drift, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
This formulation corresponds to a Wiener process with drift, in which the drift term $\drift$ only influences the expected states of nodes along the phylogeny, but does not influence the covariance structure among these nodes, which remains the same as it does in a standard Wiener process \cite{BrownianMotionHandbook}.
This allows the separation in equation \ref{ebpprior} between drift terms affecting only expectations and the evolutionary Brownian process that includes covariance among virus locations $\virus_1,\ldots,\virus_n$.

@@

Top-level priors for $1/\mdssd^2$, $\drift$, $1/\virussd^2$, and $1/\serumsd^2$ are assumed to follow diffuse $\mbox{Gamma}(a, b)$ distributions  with $a=0.001$ and $b=0.001$.
Under the full model, the posterior probability of observing virus and serum locations given immunological data is factored
\begin{equation}
	p(\viruses,\sera | \mathbf{H}) \propto p(\mathbf{H} | \viruses, \sera, \ses, \ves, \mdssd) \; 
	p(\viruses | \drift, \virussd, \tree) \;
	p(\sera | \drift, \serumsd) \; 
	p(\ses, \ves, \mdssd, \drift, \virussd, \serumsd, \tree).
\end{equation}
We sample from this posterior distribution using the MCMC procedures implemented in the software package BEAST \cite{BEAST17}.
Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus effects $\ve_i$ and serum effects $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\drift$, $\virussd$ and $\serumsd$.
For the present analysis, a two-step approach was taken to sample phylogenies, where a posterior sample of phylogenies was gathered using sequence data and then, in the cartographic analysis, trees from this set were randomly proposed and accepted following the Metropolis-Hastings algorithm \cite{Pagel04}.

@@

MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\drift$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd^2$ and phylogenetic tree $\tree$.
MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.
In all cases when drift parameter $\drift$ was included the MCMC chain mixed well and arrived at the same estimated posterior distribution from different starting points.
However, without drift parameter $\drift$, maps for A/H3N2 showed some degree of metastability, where some chains would converge on one solution and other chains would converge on a different solution.
We favor models that include $\drift$, because its inclusion, in addition to correcting most identifiability issues, yields much improved mixing of antigenic locations.
